Pangenome graph layout by Path-Guided Stochastic Gradient Descent

### 2023-07-24 manuscript/main.tex

    \subsection{Performance}
	We apply PG-SGD to the recent draft human pangenome reference \citep{Liao2023} from the Human Pangenome Reference Consortium (HPRC) to show the scalability of the algorithm with a real pangenome.
	Experiments were conducted on a cluster with 24 \texttt{Regular} nodes (32 cores / 64 threads with two AMD EPYC 7343 processors with 512 GB RAM) and 4 \texttt{HighMem} nodes (64 cores / 128 threads with two AMD EPYC 7513 processors with 2048 GB RAM).
	We downloaded pangenome graphs for each autosome (24 in total) and for the mitochondrial DNA.
	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references.
	See Supplementary Table~\ref{tab:layout} for graph statistics.
	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, PG-SGD obtains the graph layouts in 50 minutes on average, with the highest run time observed being chromosome 16 (Supplementary Table~\ref{tab:layout}).
	This is expected since chromosome 16 presents one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
	Repetitive sequences lead to graph nodes with a very high number of path traversals, which are computationally expensive to work with \citep{Guarracino2022}.
	Memory consumption is 29.66 GB of RAM on average, with the memory peak again occurring with chromosome 16, presumably due to the memory expensive path index building phase.
	Given its scalability, we even applied PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}). BandageNG\footnote{\url{https://github.com/asl/BandageNG} (last accessed Jul 2023)}, the current state-of-the-art for graph visualization, was not able to produce a layout within 7 days, hitting the wall clock time limit of the cluster.
	Overall, PG-SGD is $\approx$8X faster on average than BandageNG while using $\approx$2X less memory.
	% avg time bng 388 min, avg gb ng 66.84

    \subsection{Pangenome graphs layouts reveal biology features}
	Graph visualization is essential for understanding pangenome graphs and the genome variation they represent.
	We show how PG-SGD allows us quickly gain insight into biological data by looking at the graph layout structure.
	\FIXME{Describe a full chromosome and/or 1/a few selected loci, HLA, C4, beta-defensin locus, LPA, stuff from the human paper or PGR-TK.}

@@

	Graph visualization is key for understanding genome variations and the large-scale layouts produced by PG-SGD offer an unprecedented high-level perspective on variation in pangenomes.
	We implemented PG-SGD to generate layouts in 1 and 2 dimensions.
	These graph projections have already been employed in constructing and analyzing the first draft human pangenome reference \citep{Liao2023}, as well as in the discovery of heterologous recombination of human acrocentric chromosomes \citep{Guarracino2023}.
	Furthermore, they are applied in the creation and analysis of pangenome graphs for any species \citep{Guarracino2022, Garrison2023}.
	%The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
	%A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
	%Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
	Of note, there still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
	Our algorithm will be the foundation of new pangenome graph browsers for studying graph layouts and the genome variation they represent\footnote{\url{https://github.com/chfi/waragraph} (last accessed Jul 2023)}.

>>>>

    \subsection{Performance}
	We apply the 2D PG-SGD to the recent draft human pangenome reference \citep{Liao2023} from the Human Pangenome Reference Consortium (HPRC) to show the scalability of the algorithm with a real pangenome.
	Experiments were conducted on a cluster with 24 \texttt{Regular} nodes (32 cores / 64 threads with two AMD EPYC 7343 processors with 512 GB RAM) and 4 \texttt{HighMem} nodes (64 cores / 128 threads with two AMD EPYC 7513 processors with 2048 GB RAM).
	We downloaded pangenome graphs for each autosome (24 in total) and for the mitochondrial DNA.
	Each graph represents 90 whole human haplotypes: 44 diploid individuals plus the GRCh38 \citep{Schneider2017} and CHM13 \citep{Nurk_2021} haploid human references.
	See Supplementary Table~\ref{tab:layout} for graph statistics.
	When applied to these pangenome graphs using one \texttt{Regular} node for each calculation, PG-SGD obtains the graph layouts in 50 minutes on average, with the highest run time observed being chromosome 16 (Supplementary Table~\ref{tab:layout}).
	This is expected since chromosome 16 presents one of the highest levels of segmentally duplicated sequence among the human autosomes \citep{Martin2004}.
	Repetitive sequences lead to graph nodes with a very high number of path traversals, which are computationally expensive to work with \citep{Guarracino2022}.
	Memory consumption is 29.66 GB of RAM on average, with the memory peak again occurring with chromosome 16, presumably due to the memory expensive path index building phase.
	Given its scalability, we even applied PG-SGD to the full graph with all chromosomes together using a \texttt{HighMem} node (Supplementary Table~\ref{tab:layout}). Under one day a 2D layout was calculated. BandageNG\footnote{\url{https://github.com/asl/BandageNG} (last accessed Jul 2023)}, the current state-of-the-art for graph visualization, was not able to produce a layout within 7 days, hitting the wall clock time limit of the cluster.
	On average, PG-SGD is $\approx$8X faster than BandageNG while using $\approx$2X less memory.
	% avg time bng 388 min, avg gb ng 66.84

    \subsection{Pangenome graphs layouts reveal biology features}
	Graph visualization is essential for understanding pangenome graphs and the genome variation they represent. 
	We show how 2D PG-SGD allows us quickly gain insight into biological data by looking at the graph layout structure. 
	To demonstrate this, we provide 2D drawings of various 2D PG-SGD layouts (see Figure~\ref{fig:2d_layouts})
	In Figure~\ref{fig:sfig1} all chromosomes of the HPRC graph a drawn in 2D. 
	One can easily spot all the large scale structural variations including the hairball centromeres. 
	Focusing on the major histocompatibility complex (MHC) of chromosome 6 (Figure~\ref{fig:sfig2}), the 2D layout reveals the positions and complexity of all MHC genes (Figure~\ref{fig:sfig3}). In Figure~\ref{fig:sfig4} the C4A and C4B genes are highlighted. 
	The layout is exactly the same as the one produced by Bandage and displayed in~\citep{Guarracino2022}.

@@

	Graph visualization is key for understanding genome variations and the large-scale layouts produced by PG-SGD offer an unprecedented high-level perspective on variation in pangenomes.
	We implemented PG-SGD to generate layouts in 1D and 2D.
	These graph projections have already been employed in constructing and analyzing the first draft human pangenome reference \citep{Liao2023}, as well as in the discovery of heterologous recombination of human acrocentric chromosomes \citep{Guarracino2023}.
	Furthermore, they are applied in the creation and analysis of pangenome graphs for any species \citep{Guarracino2022, Garrison2023}.
	%The graph simplification pipeline smoothxg runs POA for each block of paths that are co-linear within each seqwish induced variation graph.
	%A prerequisite is that the graph nodes are sorted according to their occurrence in the graph's embedded paths.
	%Our 1D path-guided SGD algorithm is designed to provide this kind of sort.
	Of note, there still remains a gap in interactive and scalable solutions that merge layouts of large pangenome graph with annotation.
	Our algorithm will be the foundation of new pangenome graph browsers for studying graph layouts and the genome variation they represent\footnote{\url{https://github.com/chfi/waragraph} (last accessed Jul 2023)}.

	\FIXME{Talk about the performance evaluation here?}
